Predicting Solid State Material Platforms for Quantum Technologies

### 2022-02-27 paper_v2.tex
Updates from Overleaf

\begin{abstract}
Semiconductor materials provide a compelling platform for quantum technologies (QT), and the properties of a vast amount of materials can be found in databases containing information from both experimental and theoretical explorations. However, %scouring 
searching 
these databases to find promising candidate materials for quantum technology applications among tens of thousands of contenders is a major challenge. Therefore, we have developed a framework for the automated discovery of semiconductor host platforms for QT using material informatics and machine learning methods.

Data extraction tools have been developed for a number of materials science databases resulting in a dataset consisting of over $25.000$ materials. The information from the databases was converted into nearly $5000$ physics-informed features. Furthermore, we have developed and implemented three different approaches for the data mining process: \textit{The Ferrenti approach}, \textit{the extended Ferrenti approach}, and \textit{the empirical approach}. 
The three approaches define distinct datasets for the supervised machine learning methods logistic regression, decision trees, random forests and gradient boosting.

We find that the Ferrenti approach and the extended Ferrenti approach predict a broad range of materials as suitable for quantum technologies. The restrictions set in the empirical approach, which relies exclusively on findings from the literature, gave substantially fewer candidates with a clear distinction between suitable and unsuitable candidates when comparing the two largest eigenvalues in the covariance matrix. 
In contrast to the machine learning (ML) methods based on the Ferrenti and extended Ferrenti approaches that focused on band gap and ionic character, the ML methods from the empirical approach highlighted features related to symmetry and crystal structure, including bond length, orientation and radial distribution, as influential when labeling a material as suitable for QT.  
All four machine learning (ML) methods applied to the dataset derived using the empirical approach agreed on a set of $214$ predicted candidate materials.
All three approaches and all four machine learning methods agreed on a subset of $47$ eligible candidates %(to a probability of $>50 \ \%$) 
of $8$ elemental, $29$ binary, and $10$ tertiary compounds, and provide a basis for further material explorations towards quantum technology.

@@

The fourth science paradigm of big data driven science cover the potential of targeted search for promising material systems in which to expect QT compatible properties, see for example Ref.~\cite{Acin2018}.
Rather than searching through a host of signals for those that match our criteria, we aim to \textit{predict} which materials and signatures we should target for more detailed studies, following the framework illustrated in Figure~\ref{fig:ht-workflow}. 
This is made feasible by the availability of databases containing material properties for a wide range of different systems. In this work, the data in question are provided by bulk density functional theory (DFT) calculations to obtain the ground state properties of different elements and compounds. Combined with machine learning methods we provide a path towards precise classification of qubit candidates. The inclusion of machine learning methods follows recent trends in applications of statistical learning, data science and machine learning for scientific discoveries, see for example Refs.~\cite{deiana2021,Carleo2019}.

Here we provide a framework for the data mining and automated discovery of promising solid-state semiconductor hosts for quantum emitters and spin centers using targeted database search and machine learning methods combined with knowledge from the field. 
%\textit{En setning om discovery her??} 
Analyzing the output of the machine learning methods reveals that, given a suitable initial dataset of labeled materials for training, testing and validation, it is possible to discern the physical mechanisms that govern a material's suitability for quantum applications.  
In this framework we do not distinguish between the specific mechanism giving rise to properties such as single-photon emission and long spin coherence times (e.g., semiconductor defects or quantum dots); instead, we attempt to target all materials that may accommodate the desired characteristics.  
The methodology developed herein can be modified for other material types and application areas provided that high quality databases containing relevant theoretical and/or experimental data are available.

The approach we have developed relies on data extraction from the databases Materials Project \cite{Jain2013,Jain2018}, the Open Quantum Materials Database (OQMD) \cite{Saal2013, Kirklin2015}, JARVIS-DFT \cite{Choudhary2020}, AFLOW \cite{Curtarolo2012, Curtarolo2012a, Calderon2015} and AFLOW-ML \cite{Isayev2017}. 
The Matminer Python library for data mining \cite{Ward2018} was then used for material analysis to featurize the extracted data. An important aspect of our work is the database building and pertinent development of the datasets for machine learning methods. We have developed three different approaches to data mining: (i) \emph{the Ferrenti approach} which is similar to that proposed by \citeauthor{Ferrenti2020} \cite{Ferrenti2020}, (ii) 
%Marianne I have now changed augmented Ferrenti to extended Ferrenti 
\emph{the extended Ferrenti approach} and (iii) \emph{the empirical approach}. We find that the %former 
two first methodologies base their data extraction protocols on broad material descriptors, leading to large sets of potentially suitable candidates \cite{Mehta2019,Hastie2009}. The  
%third and 
empirical approach, on the other hand, relies on including materials with experimentally proven advantageous characteristics in the training set and therefore yields a narrower set of possible candidates.  The three training sets were then analyzed with the four supervised machine learning methods logistic regression, decision trees, random forests and gradient boosting \cite{Hastie2009,Murphy2012}, and yielded $214$ predicted candidates that were common between all approaches and machine learning methods. Importantly, the empirical approach emphasizes features like the standard deviations  of  the  radial  distribution  function  (RDF) and average bond length as critical features for suitable QT compatible materials.
%(to a $0.5$ cut-off)  
Example materials include ZnGeP$_2$, MgSe, CdS, BP, BC$_2$N, BP, Ge, GeC, InP, and InAs. 
Focused theoretical and experimental studies, using methods such as, e.g., density functional theory, photoluminescence spectroscopy and optically detected magnetic resonance, are needed to verify the predictions that quantum properties may manifest as a result of defects or nanostructures in the above listed materials.

@@

Thereafter, we apply Matminer’s \cite{Ward2018} tools to generate thousands of features from the data. 
Before applying any machine learning algorithm, the data was transformed into a numerical representation that reflects the relationship between the input and output data. This transformation is known as generating descriptors 
or features, however, we will in this work adopt the name featurization. 
The open-source library of Matminer provides many tools to featurize existing features extracted from the Materials Project. 
%MB I moved the below to methods 
%To apply Matminer's featurization tools, we extend an existing implementation by \citeauthor{Breuck2021} \cite{Breuck2021}, which were used to generate a supervised machine learning framework called the MODnet. Their implementation provides featurization for a material's composition, structure and atomic sites. However, Matminer also provides featurization tools for a material's density of states (DOS) and band structure, therefore, we extend their implementation to facilitate such featurizations. 
A schematic visualization of the featurization process in Matminer is shown in Fig.~\ref{fig:flowchart} 
and focuses on a material's composition,  structure, atomic sites, density of states and band structure. 
The $39$ features selected as material descriptors in this work are summarized and described in the Supplementary Material at \cite{supplementary}. The selection of features was kept rather wide as we do not want to make {\em ex ante} assumptions on which features best describe a solid state material platform for quantum technology.

@@

\subsection*{Data Mining}
The complete dataset consists of $25.000$ materials. A subset of these materials will be labeled into either suitable or unsuitable candidates for QT while the remainder will stay unlabeled. The labeled data are then grouped into training, test and validation sets for the machine learning methodss.

@@

\subsubsection*{The empirical approach}
The third approach differs substantially from the first two approaches in terms of material labeling, as we apply knowledge from the field (see for instance Refs.~\cite{Atatuere2018,Zhang2020,Son2020,Toth2019,Bathen2021} for an overview) to pick our promising material host candidates.  
In other words, we restrict ourselves to materials where quantum compatible properties have been either experimentally demonstrated or theoretically predicted. 
We therefore name this scheme \emph{the empirical approach}. Due to the concern of ending up with too small datasets for training, testing and validation of the machine learning methods, we will also include some  materials that are promising in the sense that they have suitable properties for accommodating deep level defects that can potentially exhibit quantum effects - even though such effects may not yet have been shown. These material choices were motivated by the criteria and predictions for QT compatible deep level defects defined by \citeauthor{Weber2010} \cite{Weber2010}. 
We have also included materials where properties like single-photon emission and spin manipulation have been observed but attributed to excitonic effects or quantum dots (formed by self-assembly or lithographic structuring) rather than being defect related.

@@

%% OH: Dirty details about suitable materials:
Entries matching the formulas C, SiC, BN, MoS$_2$, WSe$_2$ and WS$_2$ (quantum compatible characteristics have been demonstrated) were manually screened to see whether they have a matching structure to the respective candidates discussed earlier and summarized in Table~\ref{tab:qt-materials}. 
For carbon, we admit three-dimensional diamond-like structures as explicitly stated in the column tags from the Materials Project. Additionally, we find several two-dimensional structures of carbon with a large band gap ($>1.5$~eV) among the data. We add these as suitable candidates. Complex structures (e.g., C$_{28}$, C$_{48}$ and C$_{60}$) were moved to the test set in our machine learning studies. For SiC we admitted all entries which included the $2$H, $3$C, $4$H, $6$H and $15$R polytypes. Concerning BN, MoS$_2$, WSe$_2$ and WS$_2$, we only admit two-dimensional structures. For non-matching structures not mentioned so far, we move them to the test set to see if they will be predicted as suitable or not by the machine learning methods that are applied in a later stage.

@@

To illustrate the composition of the labeled data we perform a principal component analysis (PCA) \cite{Jolliffe2002}. Principal component analysis falls under the category of unsupervised learning within machine learning methods. Here we apply PCA to the labeled datasets in order to figure out if are specific features that stand out. In its standard  form, PCA relates the variance of the features with the eigenvalues of the covariance matrix \cite{Jolliffe2002,Murphy2012,Hastie2009}. We identify the two largest eigenvalues of the covariance matrix \cite{Hastie2009} of the complete initial dataset from the Materials Project, and transform the three labeled datasets according to the corresponding two eigenvectors. In essence, we reduce the dimensionality of the labeled data in order to visualize them.
The result of this procedure is displayed in the scatter plots of Figure~\ref{fig:2dscatterplotpca}. Note that some minor differences between the three panels in Fig.~\ref{fig:2dscatterplotpca} may occur due to the process of 
removing the mean and scaling to unit variance.
%centering the data. 
Red triangles represent unsuitable candidates while colored symbols (green, blue, black and yellow) represent suitable candidates. 
Due to the complexity of reducing the large amount of features down to only two, both suitable and unsuitable candidates for the Ferrenti and the extended Ferrenti approaches are largely overlapping. 
The logic behind categorizing materials in two classes (suitable and unsuitable) appears to not translate into a distinct separation, at least not in the representation of  Fig.~\ref{fig:2dscatterplotpca} for the Ferrenti and extended Ferrenti approaches.  
Hence, using these approaches for predicting QT material hosts could prove challenging for any model that would try to glean a clear-cut boundary between materials that are and are not suitable for QT. 
We expect therefore that the two Ferrenti approaches could need supplementary dimensions for further distinguishing between the materials in the two categories (suitable and unsuitable), or that it may prove challenging to find a generalized model based on the training sets that were defined according to the above discussion.

In the case of the empirical approach, on the other hand, we can discern a trend where the upper left part of the figure is dominated by suitable candidates (green, blue, yellow and black symbols), while the unsuitable candidates (red triangles) are similarly restricted to the lower right corner, albeit with some exceptions. 
Interestingly, we observe that different configurations of the famously quantum compatible materials of diamond (blue), silicon carbide (black) and silicon (yellow) are grouped together but each material class is separated in its own region. 
This indicates that the empirical approach may be capable of separating materials based on their underlying properties. Thus, devising a logical framework for the data mining process is likely important in order to obtain suitable datasets for training, testing and validation of the machine learning methods.

@@

\subsection*{Machine learning and principal component analysis}
In this work we have applied
four well-tested machine learning algorithms in order to classify specific materials as candidate systems for QT. These are logistic regression, decision trees, and ensemble methods like random forests and gradient boosting \cite{Mehta2019,Hastie2009,Murphy2012}. 
These four machine learning methods were applied to the training sets that were optimized using each of the three approaches (Ferrenti, extended Ferrenti and empirical) outlined above (recall that the labeled datasets are split into training, validation and test sets). This yielded twelve sets of materials that were predicted as suitable material hosts for QT applications.

In order to reduce the dimensionality while preserving most of the variance in the data,
%In order to deduce how the different material features are weighted by the four machine learning models when classifying materials into suitable and unsuitable candidates for QT,
we have performed a principal component analysis (PCA) \cite{Jolliffe2002} for each of the three approaches.
First, the eigenvalues of the covariance matrices of the three training sets are identified in order of decreasing value,  resulting in three distinct PCA models (as opposed to the case of Fig.~\ref{fig:2dscatterplotpca} where all panels are based on the same model).

 In the evaluation of the machine learning methods for the three different approaches we apply a $5\times 5$ stratified cross-validation \cite{Hastie2009} when searching for the optimal hyperparameter combinations. We note that all four machine learning methods have high evaluation metrics for the optimal hyperparameters. Further details on the evaluation of the machine learning methods and the results from the principal component analyses for all four machine learning methods are shown in the Supplementary Material at \cite{supplementary}.

@@

%Beneficial to use simple ML-methods for small datasets, since more advanced methods need (ALOT) more data to perform good
The three approaches define labeled data of varying size, where the smallest is from the empirical approach. For small datasets, it is proven beneficial to repeat the cross-validation, as this is a method that allows us to measure the stability of the predictions given against pertubations (i.e. few different entries) of the training data \cite{Beleites2008}.

@@

In the case of the empirical approach and in contrast to the Ferrenti based approaches, decision trees and random forests exhibit the best performance for just a few principal components, and experience a considerable degree of overfitting 
%for larger principal component values. 
when involving more principal components. Gradient boosting, in contrast to the case for the two Ferrenti based approaches, also experiences the best performance for just a few principal components. 
\begin{figure}[ht]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \input{figures/feature-importance/01-ferrenti-approach-PC-v2.tex}
        \label{fig:01-fi-e}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \input{figures/feature-importance/01-ferrenti-approach-GB-v2.tex}
        \label{fig:01-fi-d}
    \end{subfigure}%
    \hfill
    \begin{subfigure}[b]{0.45\textwidth}
        \input{figures/feature-importance/03-empirical-approach-PC-v2.tex}
        \label{fig:03-fi-e}
        \subcaption{}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \input{figures/feature-importance/03-empirical-approach-GB-v2.tex}
        \label{fig:03-fi-d}
        \subcaption{}
    \end{subfigure}
    \caption{Visualizing different parameters for the $25$ most principal components ranked in descending order by (a) the explained variance for the Ferrenti (upper panel) and empirical (lower panel) approach, and (b) the gradient boosting coefficients for the corresponding approaches. Note well that in (a) we visualize only the results of the PCA analysis while in (b) there are results of the PCA reduced training and validation sets using gradient boosting \cite{Hastie2009,xgboost2016}. The latter shows that for the empirical approach most of the physics is represented by few features (see text for an in depth discussion). The results in (b) are similar to those obtained with logistic regression, decision trees and random forests.  
    }
    \label{fig:PComponents}
\end{figure}

Fig.~\ref{fig:PComponents}(b) reveals that the single most important feature for all ML methods in the Ferrenti approach is the fifth principal component (fifth largest eigenvalue). Moreover, by selecting the highest values in this eigenvector, we find that the corresponding features originate from the DFT computed band gap of the elemental solids among the elements in the compound as calculated by Materials Agnostic Platform for Informatics and Exploration (MagPie)~\cite{magpie}. 
The second most important principal component for the Ferrenti approach is the second one. 
The second component's largest values correspond to the 
%electronegativity, ionic property, and covalence radius 
covalent radius, ionic character and packing efficiency  
among the elements in the composition. 
The data originates from elemental calculations from MagPie and are aggregated as either minimum, mean, standard deviation, or maximum. 
While the first principal component encompasses the largest explained variance, it does not provide any specific information on which features it represents.

%extended Ferrenti 
The interpretation of feature importance for the extended Ferrenti approach is substantially more difficult than in the Ferrenti approach. We find, for logistic regression and decision trees, that no feature is different than any other in the cross-validation due to a large variety of accuracy. However, we find that random forests and gradient boosting experience the fifth principal component as important. Similar to the Ferrenti approach, the corresponding features with the highest value for the fifth principal component originate from the DFT computed band gap of the elemental solids among the elements in the composition.

@@

The principal components that are deemed important by the three approaches are distinct on multiple levels. The Ferrenti and extended Ferrenti approaches place particular value on the band gap, of both the predicted compounds and the elemental solids, as well as features such as the ionic or covalent character of the bonding. Notably, these features were used as guidance in the initial selection process when forming the training sets. Thus, the machine learning methods appear to preserve the criteria imposed in the Ferrenti approaches. For the empirical approach, we did not guide the selection process in terms of specific properties, thus, other properties dominate. Rather than focusing on, e.g., the band gap, the machine learning methods appear to be informed by other characteristics more related to symmetry and crystal structure, based on the repeated appearance of bond length, orientation and radial distribution in the first principal component. In this approach, the machine learning methods appear to be inferring that the common trend among the materials that are suitable for QT is related to more complex mechanisms in the crystal structure and bonding.

\subsection*{Predicting Suitable Material Hosts for Quantum Technology} 
%In this section, 
After training the machine learning approaches using parts of the data labeled during the three data mining approaches, the ML methods were employed on the test sets. 
The number of candidates predicted by each of the four machine learning methods logistic regression, decision trees, random forests and gradient boosting is visualized in the Supplementary Material at \cite{supplementary} for each of the three approaches. 
An overview of the number of predicted suitable candidates as a function of a given machine learning approach's confidence is summarized in Table~\ref{tab:probabilites}.

@@

The findings presented here firmly establish that material informatics (from data mining via featurization to machine learning predictions) is a viable and important route to new discoveries in important fields.  Our focus has been on predicting new candidate materials to host single-photon emitters and spin centers for quantum technology applications, but the framework we have developed is suitable for other fields as well. 
Two main routes can be followed to exploit the results. One aspect is the pursuit of experimental verification of QT compatible effect in the materials predicted as suitable by the four machine learning methods. Another, perhaps even more important, is to use the features and trends identified during the data mining and prediction process to understand the distinct material characteristics that enable quantum effects to manifest, opening  thereby up for new discoveries in the field of quantum technologies.

>>>>

\begin{abstract}
Semiconductor materials provide a compelling platform for quantum technologies (QT), and the properties of a vast amount of materials can be found in databases containing information from both experimental and theoretical explorations. However, %scouring 
searching 
these databases to find promising candidate materials for quantum technology applications among tens of thousands of contenders is a major challenge. Therefore, we have developed a framework for the automated discovery of semiconductor host platforms for QT using material informatics and ML methods.

Data extraction tools have been developed for a number of materials science databases resulting in a dataset consisting of over $25.000$ materials. The information from the databases was converted into nearly $5000$ physics-informed features. Furthermore, we have developed and implemented three different approaches for the data mining process: \textit{The Ferrenti approach}, \textit{the extended Ferrenti approach}, and \textit{the empirical approach}. 
The three approaches define distinct datasets for the supervised ML methods logistic regression, decision trees, random forests and gradient boosting.

We find that the Ferrenti approach and the extended Ferrenti approach predict a broad range of materials as suitable for quantum technologies. The restrictions set in the empirical approach, which relies exclusively on findings from the literature, gave substantially fewer candidates with a clear distinction between suitable and unsuitable candidates when comparing the two largest eigenvalues in the covariance matrix. 
In contrast to the machine learning (ML) methods based on the Ferrenti and extended Ferrenti approaches that focused on band gap and ionic character, the ML methods from the empirical approach highlighted features related to symmetry and crystal structure, including bond length, orientation and radial distribution, as influential when labeling a material as suitable for QT.  
All four ML methods applied to the dataset derived using the empirical approach agreed on a set of $214$ predicted candidate materials.
All three approaches and all four ML methods agreed on a subset of $47$ eligible candidates %(to a probability of $>50 \ \%$) 
of $8$ elemental, $29$ binary, and $10$ tertiary compounds, and provide a basis for further material explorations towards quantum technology.

@@

The fourth science paradigm of big data driven science cover the potential of targeted search for promising material systems in which to expect QT compatible properties, see for example Ref.~\cite{Acin2018}.
Rather than searching through a host of signals for those that match our criteria, we aim to \textit{predict} which materials and signatures we should target for more detailed studies, following the framework illustrated in Figure~\ref{fig:ht-workflow}. 
This is made feasible by the availability of databases containing material properties for a wide range of different systems. In this work, the data in question are provided by bulk density functional theory (DFT) calculations to obtain the ground state properties of different elements and compounds. Combined with ML methods we provide a path towards precise classification of qubit candidates. The inclusion of ML methods follows recent trends in applications of statistical learning, data science and machine learning for scientific discoveries, see for example Refs.~\cite{deiana2021,Carleo2019}.

Here we provide a framework for the data mining and automated discovery of promising solid-state semiconductor hosts for quantum emitters and spin centers using targeted database search and ML methods combined with knowledge from the field. 
%\textit{En setning om discovery her??} 
Analyzing the output of the ML methods reveals that, given a suitable initial dataset of labeled materials for training, testing and validation, it is possible to discern the physical mechanisms that govern a material's suitability for quantum applications.  
In this framework we do not distinguish between the specific mechanism giving rise to properties such as single-photon emission and long spin coherence times (e.g., semiconductor defects or quantum dots); instead, we attempt to target all materials that may accommodate the desired characteristics.  
The methodology developed herein can be modified for other material types and application areas provided that high quality databases containing relevant theoretical and/or experimental data are available.

The approach we have developed relies on data extraction from the databases Materials Project \cite{Jain2013,Jain2018}, the Open Quantum Materials Database (OQMD) \cite{Saal2013, Kirklin2015}, JARVIS-DFT \cite{Choudhary2020}, AFLOW \cite{Curtarolo2012, Curtarolo2012a, Calderon2015} and AFLOW-ML \cite{Isayev2017}. 
The Matminer Python library for data mining \cite{Ward2018} was then used for material analysis to featurize the extracted data. An important aspect of our work is the database building and pertinent development of the datasets for ML methods. We have developed three different approaches to data mining: (i) \emph{the Ferrenti approach} which is similar to that proposed by \citeauthor{Ferrenti2020} \cite{Ferrenti2020}, (ii) 
%Marianne I have now changed augmented Ferrenti to extended Ferrenti 
\emph{the extended Ferrenti approach} and (iii) \emph{the empirical approach}. We find that the %former 
two first methodologies base their data extraction protocols on broad material descriptors, leading to large sets of potentially suitable candidates \cite{Mehta2019,Hastie2009}. The  
%third and 
empirical approach, on the other hand, relies on including materials with experimentally proven advantageous characteristics in the training set and therefore yields a narrower set of possible candidates.  The three training sets were then analyzed with the four supervised ML methods logistic regression, decision trees, random forests and gradient boosting \cite{Hastie2009,Murphy2012}, and yielded $214$ predicted candidates that were common between all approaches and ML methods. Importantly, the empirical approach emphasizes features like the standard deviations  of  the  radial  distribution  function  (RDF) and average bond length as critical features for suitable QT compatible materials.
%(to a $0.5$ cut-off)  
Example materials include ZnGeP$_2$, MgSe, CdS, BP, BC$_2$N, BP, Ge, GeC, InP, and InAs. 
Focused theoretical and experimental studies, using methods such as, e.g., density functional theory, photoluminescence spectroscopy and optically detected magnetic resonance, are needed to verify the predictions that quantum properties may manifest as a result of defects or nanostructures in the above listed materials.

@@

Thereafter, we apply Matminer’s \cite{Ward2018} tools to generate thousands of features from the data. 
Before applying any ML algorithm, the data was transformed into a numerical representation that reflects the relationship between the input and output data. This transformation is known as generating descriptors 
or features, however, we will in this work adopt the name featurization. 
The open-source library of Matminer provides many tools to featurize existing features extracted from the Materials Project. 
%MB I moved the below to methods 
%To apply Matminer's featurization tools, we extend an existing implementation by \citeauthor{Breuck2021} \cite{Breuck2021}, which were used to generate a supervised machine learning framework called the MODnet. Their implementation provides featurization for a material's composition, structure and atomic sites. However, Matminer also provides featurization tools for a material's density of states (DOS) and band structure, therefore, we extend their implementation to facilitate such featurizations. 
A schematic visualization of the featurization process in Matminer is shown in Fig.~\ref{fig:flowchart} 
and focuses on a material's composition,  structure, atomic sites, density of states and band structure. 
The $39$ features selected as material descriptors in this work are summarized and described in the Supplementary Material at \cite{supplementary}. The selection of features was kept rather wide as we do not want to make {\em ex ante} assumptions on which features best describe a solid state material platform for quantum technology.

@@

\subsection*{Data Mining}
The complete dataset consists of $25.000$ materials. A subset of these materials will be labeled into either suitable or unsuitable candidates for QT while the remainder will stay unlabeled. The labeled data are then grouped into training, test and validation sets for the ML methods.

@@

\subsubsection*{The empirical approach}
The third approach differs substantially from the first two approaches in terms of material labeling, as we apply knowledge from the field (see for instance Refs.~\cite{Atatuere2018,Zhang2020,Son2020,Toth2019,Bathen2021} for an overview) to pick our promising material host candidates.  
In other words, we restrict ourselves to materials where quantum compatible properties have been either experimentally demonstrated or theoretically predicted. 
We therefore name this scheme \emph{the empirical approach}. Due to the concern of ending up with too small datasets for training, testing and validation of the ML methods, we will also include some  materials that are promising in the sense that they have suitable properties for accommodating deep level defects that can potentially exhibit quantum effects - even though such effects may not yet have been shown. These material choices were motivated by the criteria and predictions for QT compatible deep level defects defined by \citeauthor{Weber2010} \cite{Weber2010}. 
We have also included materials where properties like single-photon emission and spin manipulation have been observed but attributed to excitonic effects or quantum dots (formed by self-assembly or lithographic structuring) rather than being defect related.

@@

%% OH: Dirty details about suitable materials:
Entries matching the formulas C, SiC, BN, MoS$_2$, WSe$_2$ and WS$_2$ (quantum compatible characteristics have been demonstrated) were manually screened to see whether they have a matching structure to the respective candidates discussed earlier and summarized in Table~\ref{tab:qt-materials}. 
For carbon, we admit three-dimensional diamond-like structures as explicitly stated in the column tags from the Materials Project. Additionally, we find several two-dimensional structures of carbon with a large band gap ($>1.5$~eV) among the data. We add these as suitable candidates. Complex structures (e.g., C$_{28}$, C$_{48}$ and C$_{60}$) were moved to the test set in our machine learning studies. For SiC we admitted all entries which included the $2$H, $3$C, $4$H, $6$H and $15$R polytypes. Concerning BN, MoS$_2$, WSe$_2$ and WS$_2$, we only admit two-dimensional structures. For non-matching structures not mentioned so far, we move them to the test set to see if they will be predicted as suitable or not by the ML methods that are applied in a later stage.

@@

To illustrate the composition of the labeled data we perform a principal component analysis (PCA) \cite{Jolliffe2002}. Principal component analysis falls under the category of unsupervised learning within ML methods. Here we apply PCA to the labeled datasets in order to figure out if there are specific features that stand out. In its standard  form, PCA relates the variance of the features with the eigenvalues of the covariance matrix \cite{Jolliffe2002,Murphy2012,Hastie2009}. We identify the two largest eigenvalues of the covariance matrix \cite{Hastie2009} of the complete initial dataset from the Materials Project, and transform the three labeled datasets according to the corresponding two eigenvectors. In essence, we reduce the dimensionality of the labeled data in order to visualize them.
The result of this procedure is displayed in the scatter plots of Fig.~\ref{fig:2dscatterplotpca}. Note that some minor differences between the three panels in Fig.~\ref{fig:2dscatterplotpca} may occur due to the process of 
removing the mean and scaling to unit variance.
%centering the data. 
Red triangles represent unsuitable candidates while colored symbols (green, blue, black and yellow) represent suitable candidates. 
Due to the complexity of reducing the large amount of features down to only two, both suitable and unsuitable candidates for the Ferrenti and the extended Ferrenti approaches are largely overlapping. 
The logic behind categorizing materials in two classes (suitable and unsuitable) appears to not translate into a distinct separation, at least not in the representation of  Fig.~\ref{fig:2dscatterplotpca} for the Ferrenti and extended Ferrenti approaches.  
Hence, using these approaches for predicting QT material hosts could prove challenging for any model that would try to glean a clear-cut boundary between materials that are and are not suitable for QT. 
We expect therefore that the two Ferrenti approaches could need supplementary dimensions for further distinguishing between the materials in the two categories (suitable and unsuitable), or that it may prove challenging to find a generalized model based on the training sets that were defined according to the above discussion.

In the case of the empirical approach, on the other hand, we can discern a trend where the upper left part of the figure is dominated by suitable candidates (green, blue, yellow and black symbols), while the unsuitable candidates (red triangles) are similarly restricted to the lower right corner, albeit with some exceptions. 
Interestingly, we observe that different configurations of the famously quantum compatible materials of diamond (blue), silicon carbide (black) and silicon (yellow) are grouped together but each material class is separated in its own region. 
This indicates that the empirical approach may be capable of separating materials based on their underlying properties. Thus, devising a logical framework for the data mining process is likely important in order to obtain suitable datasets for training, testing and validation of the ML methods.

@@

\subsection*{Machine learning and principal component analysis}
In this work we have applied
four well-tested machine learning algorithms in order to classify specific materials as candidate systems for QT. These are logistic regression, decision trees, and ensemble methods like random forests and gradient boosting \cite{Mehta2019,Hastie2009,Murphy2012}. 
These four ML methods were applied to the training sets that were optimized using each of the three approaches (Ferrenti, extended Ferrenti and empirical) outlined above (recall that the labeled datasets are split into training, validation and test sets). This yielded twelve sets of materials that were predicted as suitable material hosts for QT applications.

In order to reduce the dimensionality while preserving most of the variance in the data,
%In order to deduce how the different material features are weighted by the four machine learning models when classifying materials into suitable and unsuitable candidates for QT,
we have performed a principal component analysis (PCA) \cite{Jolliffe2002} for each of the three approaches (Ferrenti, extended Ferrenti and empirical).
First, the eigenvalues of the covariance matrices of the three training sets are identified in order of decreasing value,  resulting in three distinct PCA models (as opposed to the case of Fig.~\ref{fig:2dscatterplotpca} where all panels are based on the same model).

 In the evaluation of the ML methods for the three different approaches we apply a $5\times 5$ stratified cross-validation \cite{Hastie2009} when searching for the optimal hyperparameter combinations. We note that all four ML methods have high evaluation metrics for the optimal hyperparameters. Further details on the evaluation of the ML methods and the results from the principal component analyses for all four methods are shown in the Supplementary Material at \cite{supplementary}.

@@

%Beneficial to use simple ML-methods for small datasets, since more advanced methods need (ALOT) more data to perform good
The three data set approaches (Ferrenti, extended Ferrenti and empirical) define labeled data of varying size, where the smallest is from the empirical approach. For small datasets, it is proven beneficial to repeat the cross-validation analysis \cite{Hastie2009}, as this is a method that allows us to measure the stability of the predictions  against perturbations (i.e. few different entries) in the training data \cite{Beleites2008}.

@@

In the case of the empirical approach and in contrast to the Ferrenti based approaches, decision trees and random forests exhibit the best performance for just a few principal components, and experience a considerable degree of overfitting 
%for larger principal component values. 
when involving more principal components. Gradient boosting, in contrast to the case for the two Ferrenti  approaches, also experiences the best performance for just a few principal components. 
\begin{figure}[ht]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \input{figures/feature-importance/01-ferrenti-approach-PC-v2.tex}
        \label{fig:01-fi-e}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \input{figures/feature-importance/01-ferrenti-approach-GB-v2.tex}
        \label{fig:01-fi-d}
    \end{subfigure}%
    \hfill
    \begin{subfigure}[b]{0.45\textwidth}
        \input{figures/feature-importance/03-empirical-approach-PC-v2.tex}
        \label{fig:03-fi-e}
        \subcaption{}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \input{figures/feature-importance/03-empirical-approach-GB-v2.tex}
        \label{fig:03-fi-d}
        \subcaption{}
    \end{subfigure}
    \caption{Visualizing different parameters for the $25$ most principal components ranked in descending order by (a) the explained variance for the Ferrenti (upper panel) and empirical (lower panel) approach, and (b) the gradient boosting coefficients for the corresponding approaches. Note well that in (a) we visualize only the results of the PCA analysis while in (b) there are results of the PCA reduced training and validation sets using gradient boosting \cite{Hastie2009,xgboost2016}. The latter shows that for the empirical approach most of the physics is represented by few features (see text for an in depth discussion). The results in (b) are similar to those obtained with logistic regression, decision trees and random forests.  
    }
    \label{fig:PComponents}
\end{figure}

Figure~\ref{fig:PComponents}(b) reveals that the single most important feature for all ML methods in the Ferrenti approach is the fifth principal component (fifth largest eigenvalue). Moreover, by selecting the highest values in this eigenvector, we find that the corresponding features originate from the DFT computed band gap of the elemental solids among the elements in the compound as calculated by Materials Agnostic Platform for Informatics and Exploration (MagPie)~\cite{magpie}. 
The second most important principal component for the Ferrenti approach receives significant contributions from the  
%electronegativity, ionic property, and covalence radius 
covalent radius, the ionic character and the packing efficiency  
among the elements in the composition. 
The data originates from elemental calculations from MagPie and are aggregated as either minimum, mean, standard deviation, or maximum. 
While the first principal component encompasses the largest explained variance, it does not provide a clear and  specific information on which features it represents.

%extended Ferrenti 
The interpretation of feature importance for the extended Ferrenti approach is substantially more difficult than in the Ferrenti approach. We find, for logistic regression and decision trees that no feature is different than any other in the cross-validation, mainly due to a large variety of accuracy. However, we find that random forests and gradient boosting show the importance of  fifth principal component. Similar to the Ferrenti approach, the corresponding features with the highest value for the fifth principal component originate from the DFT computed band gap of the elemental solids among the elements in the composition.

@@

The principal components that are deemed important by the three approaches are distinct on multiple levels. The Ferrenti and extended Ferrenti approaches place particular value on the band gap, of both the predicted compounds and the elemental solids, as well as features such as the ionic or covalent character of the bonding. Notably, these features were used as guidance in the initial selection process when forming the training sets. Thus, the ML methods appear to preserve the criteria imposed in the Ferrenti approaches. For the empirical approach, we did not guide the selection process in terms of specific properties, thus, other properties dominate. Rather than focusing on, e.g., the band gap, the ML methods appear to be informed by other characteristics more related to symmetry and crystal structure, based on the repeated appearance of bond length, orientation and radial distribution in the first principal component. In this approach, the ML methods appear to be inferring that the common trend among the materials that are suitable for QT is related to more complex mechanisms in the crystal structure and bonding.

\subsection*{Predicting Suitable Material Hosts for Quantum Technology} 
%In this section, 
After training the machine learning approaches using parts of the data labeled during the three data mining approaches, the ML methods were employed on the test sets. 
The number of candidates predicted by each of the four ML methods logistic regression, decision trees, random forests and gradient boosting is visualized in the Supplementary Material at \cite{supplementary} for each of the three approaches. 
An overview of the number of predicted suitable candidates as a function of a given machine learning approach's confidence is summarized in Table~\ref{tab:probabilites}.

@@

The findings presented here firmly establish that material informatics (from data mining via featurization to machine learning predictions) is a viable and important route to new discoveries in important fields.  Our focus has been on predicting new candidate materials to host single-photon emitters and spin centers for quantum technology applications, but the framework we have developed is suitable for other fields as well. 
Two main routes can be followed to exploit the results. One aspect is the pursuit of experimental verification of QT compatible effect in the materials predicted as suitable by the four ML methods. Another, perhaps even more important, is to use the features and trends identified during the data mining and prediction process to understand the distinct material characteristics that enable quantum effects to manifest, opening  thereby up for new discoveries in the field of quantum technologies.
